### 2021-04-13 ch01_introduction/introduction.tex
Fixed formatting of long numbers and added missing page information for refs

Climate change is one of the greatest challenges for humankind today. The
warming of the climate system is \enquote{unequivocal}, and \enquote{many of
  the observed changes are unprecedented over decades to millennia}
\autocite{IPCC2014}. The changing climate increases the \enquote{likelihood of
  severe, pervasive and irreversible impacts for people, species and
  ecosystems} with \enquote{mostly negative impacts for biodiversity, ecosystem
services and economic development}, and amplifies \enquote{risks for
livelihoods and for food and human security} \autocite{IPCC2014}. Potential
drivers for climate change are all natural and anthropogenic substances and
processes that may alter the Earth's energy budget. Human influence on the
climate system is clear: Ever-increasing emissions of \acp{GHG} since the end
of the pre-industrial era largely driven by economic and population growth
led to atmospheric concentrations of \ac{CO2}, \ac{CH4}, and \ac{N2O} that
are \enquote{unprecedented in at least the last 800000 years}
\autocite{IPCC2014}. The effects of these \ac{GHG} emissions and other
anthropogenic drivers have been \enquote{detected throughout the climate
system and are extremely likely to have been the dominant cause of the
observed warming since the mid-\nth{20} century} \autocite{IPCC2014}. The
greenhouse effect is based on the optical properties of the \acp{GHG}: While
the atmosphere is mostly transparent for the incoming solar (shortwave)
radiation, the outgoing infrared (longwave) radiation that is reflected from
the planet's surface is partly absorbed by \ac{GHG} molecules through
excitation of their corresponding vibrational modes and re-emitted
isotropically. This process traps energy near the surface and leads to a
warming of the Earth's surface and the lower atmosphere. The individual
impact of the different drivers of climate change is measured with the
so-called \emph{radiative forcing}, which quantifies the change in energy
fluxes caused by changes in these drivers relative to pre-industrial
conditions \autocite{IPCC2013}. The total radiative forcing is positive
(which corresponds to a warming) and its largest contribution is caused by
the increase in the atmospheric concentration of \ac{CO2} due to
anthropogenic fossil fuel emissions since the mid-\nth{18} century
\autocite{IPCC2014}. Apart from their physical warming effect on the climate,
carbon-based \acp{GHG} like \ac{CO2} also directly influence the global
carbon cycle, an important biogeochemical cycle of the Earth. A crucial flux
of the carbon cycle is the \emph{\ac{GPP}}, which describes the carbon uptake
of the terrestrial biosphere due to photosynthesis. Since this land carbon
uptake absorbs about $30 \unit{\%}$ of the anthropogenic \ac{CO2} emissions
in today's climate \autocite{Friedlingstein2020}, this process substantially
slows down global warming and directly contributes to the magnitude of
climate sensitivity. Other important anthropogenic drivers of climate change
are the emission of aerosols and land-use/land-cover changes. Apart from
anthropogenic drivers, there are also natural processes that impact the
climate system like changes in solar activity or volcanic eruptions. However,
there is clear evidence that these natural drivers alone cannot explain the
observed climate change \autocite{Haustein2017}.

>>>>

Climate change is one of the greatest challenges for humankind today. The
warming of the climate system is \enquote{unequivocal}, and \enquote{many of
  the observed changes are unprecedented over decades to millennia}
\autocite{IPCC2014}. The changing climate increases the \enquote{likelihood of
  severe, pervasive and irreversible impacts for people, species and
  ecosystems} with \enquote{mostly negative impacts for biodiversity, ecosystem
services and economic development}, and amplifies \enquote{risks for
livelihoods and for food and human security} \autocite{IPCC2014}. Potential
drivers for climate change are all natural and anthropogenic substances and
processes that may alter the Earth's energy budget. Human influence on the
climate system is clear: Ever-increasing emissions of \acp{GHG} since the end
of the pre-industrial era largely driven by economic and population growth
led to atmospheric concentrations of \ac{CO2}, \ac{CH4}, and \ac{N2O} that
are \enquote{unprecedented in at least the last 800,000 years}
\autocite{IPCC2014}. The effects of these \ac{GHG} emissions and other
anthropogenic drivers have been \enquote{detected throughout the climate
system and are extremely likely to have been the dominant cause of the
observed warming since the mid-\nth{20} century} \autocite{IPCC2014}. The
greenhouse effect is based on the optical properties of the \acp{GHG}: While
the atmosphere is mostly transparent for the incoming solar (shortwave)
radiation, the outgoing infrared (longwave) radiation that is reflected from
the planet's surface is partly absorbed by \ac{GHG} molecules through
excitation of their corresponding vibrational modes and re-emitted
isotropically. This process traps energy near the surface and leads to a
warming of the Earth's surface and the lower atmosphere. The individual
impact of the different drivers of climate change is measured with the
so-called \emph{radiative forcing}, which quantifies the change in energy
fluxes caused by changes in these drivers relative to pre-industrial
conditions \autocite{IPCC2013}. The total radiative forcing is positive
(which corresponds to a warming) and its largest contribution is caused by
the increase in the atmospheric concentration of \ac{CO2} due to
anthropogenic fossil fuel emissions since the mid-\nth{18} century
\autocite{IPCC2014}. Apart from their physical warming effect on the climate,
carbon-based \acp{GHG} like \ac{CO2} also directly influence the global
carbon cycle, an important biogeochemical cycle of the Earth. A crucial flux
of the carbon cycle is the \emph{\ac{GPP}}, which describes the carbon uptake
of the terrestrial biosphere due to photosynthesis. Since this land carbon
uptake absorbs about $30 \unit{\%}$ of the anthropogenic \ac{CO2} emissions
in today's climate \autocite{Friedlingstein2020}, this process substantially
slows down global warming and directly contributes to the magnitude of
climate sensitivity. Other important anthropogenic drivers of climate change
are the emission of aerosols and land-use/land-cover changes. Apart from
anthropogenic drivers, there are also natural processes that impact the
climate system like changes in solar activity or volcanic eruptions. However,
there is clear evidence that these natural drivers alone cannot explain the
observed climate change \autocite{Haustein2017}.

### 2021-04-13 ch02_scientific_background/scientific_background.tex
Fixed formatting of long numbers and added missing page information for refs

A schematic overview of the global carbon cycle is shown in
\cref{fig:02:carbon_cycle_schematic}. To quantify the carbon cycle, common
units are \ac{ppm} for the atmospheric trace gas concentrations (dry-air mole
fraction) and \ac{GtC} or $\si{\GtCyr{}}$ for the reservoir masses or exchange
fluxes, respectively. The carbon exchange processes between the different
carbon reservoirs run on a wide range of time scales. Conceptually, one can
distinguish between two domains of the global carbon cycle: a slow and a fast
domain. The slow domain with turnover times (reservoir mass of carbon divided
by exchange flux) of more than 10000 years consists of the large carbon stores
in rocks and sediments which are connected to the rapid domain of the carbon
cycle through volcanic emissions of \ac{CO2}, chemical weathering, erosion, and
sediment formation on the seafloor. These natural exchange fluxes between the
slow and the fast domain are comparatively small ($< 0.3 \unit{\GtCyr{}}$) and
can be assumed as approximately constant in time over the last few centuries
\autocite{Ciais2013}.

The fast domain of the global carbon cycle consists of three main carbon
reservoirs: the atmosphere, the terrestrial biosphere, and the ocean. In the
atmosphere, carbon is mainly stored in trace gases, with \ac{CO2} as the major
component with a current (2019) concentration of about $410 \unit{ppm}$
\autocite{Friedlingstein2020}. Additional contributors to the atmospheric
carbon content are the trace gas \ac{CH4}, the trace gas \ac{CO}, hydrocarbons,
black carbon aerosols, and organic compounds \autocite{Ciais2013}. Carbon in
the terrestrial biosphere is mainly stored as organic compounds, with about
$\rangeunit{450}{650}{GtC}$ in the living vegetation biomass,
$\rangeunit{1500}{2400}{GtC}$ in dead organic matter in litter and soils, and
about $1700 \unit{GtC}$ in permafrost soils \autocite{Ciais2013}. The main
component of the oceanic carbon reservoir is dissolved inorganic carbon
(carbonic acid, bicarbonate ions, and carbonate ions) with about $38000
\unit{GtC}$. Further carbon is stored as dissolved organic carbon (about $700
\unit{GtC}$), in surface sediments (about $1750 \unit{GtC}$), and in marine
biota (about $3 \unit{GtC}$, predominantly phytoplankton and other
microorganisms) \autocite{Ciais2013, Friedlingstein2020}.

>>>>

A schematic overview of the global carbon cycle is shown in
\cref{fig:02:carbon_cycle_schematic}. To quantify the carbon cycle, common
units are \ac{ppm} for the atmospheric trace gas concentrations (dry-air mole
fraction) and \ac{GtC} or $\si{\GtCyr{}}$ for the reservoir masses or exchange
fluxes, respectively. The carbon exchange processes between the different
carbon reservoirs run on a wide range of time scales. Conceptually, one can
distinguish between two domains of the global carbon cycle: a slow and a fast
domain. The slow domain with turnover times (reservoir mass of carbon divided
by exchange flux) of more than 10,000 years consists of the large carbon stores
in rocks and sediments which are connected to the rapid domain of the carbon
cycle through volcanic emissions of \ac{CO2}, chemical weathering, erosion, and
sediment formation on the seafloor. These natural exchange fluxes between the
slow and the fast domain are comparatively small ($< 0.3 \unit{\GtCyr{}}$) and
can be assumed as approximately constant in time over the last few centuries
\autocite{Ciais2013}.

The fast domain of the global carbon cycle consists of three main carbon
reservoirs: the atmosphere, the terrestrial biosphere, and the ocean. In the
atmosphere, carbon is mainly stored in trace gases, with \ac{CO2} as the major
component with a current (2019) concentration of about $410 \unit{ppm}$
\autocite{Friedlingstein2020}. Additional contributors to the atmospheric
carbon content are the trace gas \ac{CH4}, the trace gas \ac{CO}, hydrocarbons,
black carbon aerosols, and organic compounds \autocite{Ciais2013}. Carbon in
the terrestrial biosphere is mainly stored as organic compounds, with about
$\rangeunit{450}{650}{GtC}$ in the living vegetation biomass,
$\rangeunit{1500}{2400}{GtC}$ in dead organic matter in litter and soils, and
about $1700 \unit{GtC}$ in permafrost soils \autocite{Ciais2013}. The main
component of the oceanic carbon reservoir is dissolved inorganic carbon
(carbonic acid, bicarbonate ions, and carbonate ions) with about
$\mathtxt{38,000} \unit{GtC}$. Further carbon is stored as dissolved organic
carbon (about $700 \unit{GtC}$), in surface sediments (about $1750
\unit{GtC}$), and in marine biota (about $3 \unit{GtC}$, predominantly
phytoplankton and other microorganisms) \autocite{Ciais2013,
  Friedlingstein2020}.

### 2021-04-13 ch03_esmvaltool/esmvaltool.tex
Fixed formatting of long numbers and added missing page information for refs

As of \TheMonth{}, 41780 lines of code have been added and 21713 lines of code
have been removed from the GitHub repository of \ac{ESMValCore} by the author
of this thesis. Apart from general improvements of the code base, these changes
and additions mainly include new preprocessor functions that can be applied to
the input data and derivation scripts for new non-\ac{CMOR} variables.
\Cref{tab:03:changes_to_esmvalcore} shows a summary of these main
contributions, which are partly published in the scientific documentation of
\ac{ESMValCore} \autocite{Righi2020}.

@@

In addition to the analyses presented in
\cref{ch:04:papers_ecs_tcr_assessment,ch:05:paper_ecs,ch:06:paper_gpp}, several
more recipes and \acs{CMOR}ization scripts for observational data have been
added to the main \ac{ESMValTool} repository on GitHub. A summary of these is
given in \cref{tab:03:changes_to_esmvaltool}. Moreover, multiple minor changes
and additions have been implemented to improve the code base and/or to fix
bugs. In total, the author of this study has added 98662 lines of code and
removed 49823 lines of code from the \ac{ESMValTool} repository (as of
\TheMonth{}). Parts of these implementations are already published in the
scientific documentation of the \ac{ESMValTool}, which covers large-scale
diagnostics \autocite{Eyring2020}, diagnostics for emergent constraints and
future projections \autocite{Lauer2020}, and diagnostics for extreme events and
regional evaluation \autocite{Weigel2020}.

>>>>

As of \TheMonth{}, 41,780 lines of code have been added and 21,713 lines of
code have been removed from the GitHub repository of \ac{ESMValCore} by the
author of this thesis. Apart from general improvements of the code base, these
changes and additions mainly include new preprocessor functions that can be
applied to the input data and derivation scripts for new non-\ac{CMOR}
variables. \Cref{tab:03:changes_to_esmvalcore} shows a summary of these main
contributions, which are partly published in the scientific documentation of
\ac{ESMValCore} \autocite{Righi2020}.

@@

In addition to the analyses presented in
\cref{ch:04:papers_ecs_tcr_assessment,ch:05:paper_ecs,ch:06:paper_gpp}, several
more recipes and \acs{CMOR}ization scripts for observational data have been
added to the main \ac{ESMValTool} repository on GitHub. A summary of these is
given in \cref{tab:03:changes_to_esmvaltool}. Moreover, multiple minor changes
and additions have been implemented to improve the code base and/or to fix
bugs. In total, the author of this study has added 98,662 lines of code and
removed 49,823 lines of code from the \ac{ESMValTool} repository (as of
\TheMonth{}). Parts of these implementations are already published in the
scientific documentation of the \ac{ESMValTool}, which covers large-scale
diagnostics \autocite{Eyring2020}, diagnostics for emergent constraints and
future projections \autocite{Lauer2020}, and diagnostics for extreme events and
regional evaluation \autocite{Weigel2020}.

### 2021-04-13 ch06_paper_gpp/paper_gpp.tex
Fixed formatting of long numbers and added missing page information for refs

Our \ac{GBRT} approach (Step 2) is mathematically similar to \ac{MDER}
\autocite{Karpechko2013, Senftleben2020, Wenzel2016a}: We establish a
relationship between process-oriented, physically relevant diagnostics and
future projections and then utilize this to project today's observed conditions
into the future. We emphasize that the exact nature of this relationship is
strongly dependent on the climate model ensemble considered. Although derived
empirically, we argue that this relationship can still be used to extract
further information from the climate model ensemble for two reasons: First, we
consider process-oriented variables which are physically linked to \ac{GPP},
and second, we ensure its statistical robustness. The latter is achieved by
considering gridded climatological data instead of global or regional means.
The number of points establishing the relation is dramatically increased from
about 50 points for classical emergent constraints to 237852 points (Step 2a)
and 16503 (Step 2b) in our approach. We validate the relationship by extensive
out-of-sample testing using a leave-one-model-out \ac{CV} approach and randomly
excluding parts of the data prior to training. Furthermore, the large number of
points in combination with the non-linear \ac{GBRT} model allows us to exploit
more information than classical emergent constraints and include non-linear
multivariate relations in the statistical model, including the information
encoded in spatial variation. The prediction phase of our approach (feeding
observation-based data into the trained \ac{GBRT} model) can be interpreted as
an implicit performance weighting: The \ac{GBRT} model creates predictions
based on today's conditions (in form of variables that are physically connected
to \ac{GPP}). However, unlike other performance weighting techniques
\autocite{Knutti2017a, Sanderson2017}, we do not assign constant weights to the
different climate models. On the contrary, since the algorithm works on
grid-cell level, it is able to adapt to regional climate characteristics and
thus applies an implicit local and climate-dependent weighting. Moreover, the
algorithm adds another level of weighting by assigning a relative relevance
score to the different predictors. We emphasize that this is only an implicit
weighting since it is not possible to extract specific values for the
individual weights due to the complex structure of the \ac{GBRT} model.

>>>>

Our \ac{GBRT} approach (Step 2) is mathematically similar to \ac{MDER}
\autocite{Karpechko2013, Senftleben2020, Wenzel2016a}: We establish a
relationship between process-oriented, physically relevant diagnostics and
future projections and then utilize this to project today's observed conditions
into the future. We emphasize that the exact nature of this relationship is
strongly dependent on the climate model ensemble considered. Although derived
empirically, we argue that this relationship can still be used to extract
further information from the climate model ensemble for two reasons: First, we
consider process-oriented variables which are physically linked to \ac{GPP},
and second, we ensure its statistical robustness. The latter is achieved by
considering gridded climatological data instead of global or regional means.
The number of points establishing the relation is dramatically increased from
about 50 points for classical emergent constraints to 237,852 points (Step 2a)
and 16,503 (Step 2b) in our approach. We validate the relationship by extensive
out-of-sample testing using a leave-one-model-out \ac{CV} approach and randomly
excluding parts of the data prior to training. Furthermore, the large number of
points in combination with the non-linear \ac{GBRT} model allows us to exploit
more information than classical emergent constraints and include non-linear
multivariate relations in the statistical model, including the information
encoded in spatial variation. The prediction phase of our approach (feeding
observation-based data into the trained \ac{GBRT} model) can be interpreted as
an implicit performance weighting: The \ac{GBRT} model creates predictions
based on today's conditions (in form of variables that are physically connected
to \ac{GPP}). However, unlike other performance weighting techniques
\autocite{Knutti2017a, Sanderson2017}, we do not assign constant weights to the
different climate models. On the contrary, since the algorithm works on
grid-cell level, it is able to adapt to regional climate characteristics and
thus applies an implicit local and climate-dependent weighting. Moreover, the
algorithm adds another level of weighting by assigning a relative relevance
score to the different predictors. We emphasize that this is only an implicit
weighting since it is not possible to extract specific values for the
individual weights due to the complex structure of the \ac{GBRT} model.

### 2021-04-13 end_appendix/appendix.tex
Added acknowledgements
Fixed formatting of long numbers and added missing page information for refs

\begin{table}[!h]
  \centering
  \begin{tabulary}{\columnwidth}{>{\em}L L}
    \toprule
    \ignorecolumntype{l}{Variable short name} & Description \\
    \midrule
    cl & Cloud area fraction \\
    clt & Total cloud area fraction \\
    hur & Relative humidity \\
    hus & Specific humidity \\
    pr & \Acl{PR} \\
    rsdt & \Acf{TOA} incoming shortwave radiation \\
    rsut & \acs{TOA} outgoing shortwave radiation \\
    rsutcs & Clear-sky \acs{TOA} outgoing shortwave radiation \\
    ta & Air temperature \\
    tas & Surface air temperature \\
    tos & Sea surface temperature \\
    ts & Surface temperature \\
    va & Northward wind speed \\
    wap & Vertical velocity \\
    \bottomrule
  \end{tabulary}
  \caption[
    Overview of the variables used in \cref{ch:05:paper_ecs}.
  ]{
    Overview of the variables used in \cref{ch:05:paper_ecs}. More details are
    given in \cref{tab:05:overview_emergent_constraints}, which lists the
    variables used for each emergent constraint.
  }
  \label{tab:app:a:overview_variables}
\end{table}

@@

The raw monthly mean output of every participating climate model and
observation-driven dataset is regridded to a $\ang{2} \times \ang{2}$ grid and
masked with a common mask to remove all oceans and Antarctica. For Step 2a
(target variable: absolute \acf{GPP} at the end of the \nth{21} century),
monthly climatologies are calculated for every dataset by averaging over all
available years for every month. For Step 2b (target variable: fractional
\ac{GPP} change over the \nth{21} century), temporal means are calculated for
every dataset by averaging over the full time dimension. In addition, values
greater than $300 \unit{\%}$ in the target variable in Step 2b (fractional
\ac{GPP} change over the \nth{21} century) are masked to avoid numerical
inconsistencies caused by the division of small numbers in the derivation of
the target variable. In the next step, the multi-dimensional data is flattened
and all the training data from the different climate models is stacked into a
single large training array. Finally, to account for the varying magnitudes of
the different predictors, all of them are linearly scaled by their respective
means and standard deviations so that they have a mean of zero and unit
variance. In total, 237852 (16503) training data points, 79284 (5501) hold-out
test data points, and 46344 (3727) points for the prediction are used in the
machine learning model for Step 2a (Step 2b).

>>>>

\begin{table}[!h]
  \centering
  \begin{tabulary}{\columnwidth}{>{\em}L L}
    \toprule
    \ignorecolumntype{l}{Variable short name} & Description \\
    \midrule
    cl & Cloud area fraction \\
    clt & Total cloud area fraction \\
    hur & Relative humidity \\
    hus & Specific humidity \\
    pr & \Acl{PR} \\
    rsdt & \Acf{TOA} incoming shortwave radiation \\
    rsut & \acs{TOA} outgoing shortwave radiation \\
    rsutcs & Clear-sky \acs{TOA} outgoing shortwave radiation \\
    ta & Air temperature \\
    tas & Surface air temperature \\
    tos & Sea surface temperature \\
    ts & Surface temperature \\
    va & Northward wind speed \\
    wap & Vertical velocity \\
    \bottomrule
  \end{tabulary}
  \caption[
    Overview of the variables used in \cref{ch:05:paper_ecs}.
  ]{
    Overview of the variables used in \cref{ch:05:paper_ecs}. More details are
    given in \cref{tab:05:overview_emergent_constraints}, which lists the
    individual variables used for each emergent constraint.
  }
  \label{tab:app:a:overview_variables}
\end{table}

@@

The raw monthly mean output of every participating climate model and
observation-driven dataset is regridded to a $\ang{2} \times \ang{2}$ grid and
masked with a common mask to remove all oceans and Antarctica. For Step 2a
(target variable: absolute \acf{GPP} at the end of the \nth{21} century),
monthly climatologies are calculated for every dataset by averaging over all
available years for every month. For Step 2b (target variable: fractional
\ac{GPP} change over the \nth{21} century), temporal means are calculated for
every dataset by averaging over the full time dimension. In addition, values
greater than $300 \unit{\%}$ in the target variable in Step 2b (fractional
\ac{GPP} change over the \nth{21} century) are masked to avoid numerical
inconsistencies caused by the division of small numbers in the derivation of
the target variable. In the next step, the multi-dimensional data is flattened
and all the training data from the different climate models is stacked into a
single large training array. Finally, to account for the varying magnitudes of
the different predictors, all of them are linearly scaled by their respective
means and standard deviations so that they have a mean of zero and unit
variance. In total, 237,852 (16,503) training data points, 79,284 (5501)
hold-out test data points, and 46,344 (3727) points for the prediction are used
in the machine learning model for Step 2a (Step 2b).
